Parameterized convection, stratiform clouds and resolution sensitivity in the Community Atmosphere Model

### 2019-09-10 resolution.tex
edits

@@

The large increase in magnitude of the upward and downward vertical velocities with resolution may be expected to impact the behavior of other model components. Curiously, there is an excellent negative correlation (Pearson's R-value = 0.99, N = 27) between the global mean, climatological $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ and a measure of the activity of the \cite{ZM1995AO} deep convection scheme (referred to as the {\em{ZM scheme}} hereafter), global mean, climatological $FREQZM$ (Figure~\ref{fig:corr}). At any grid-point and time-step, $FREQZM$ is a binary variable: 1 if the ZM scheme is active, 0 if it is not. Time mean $FREQZM$ is therefore the fraction of the model time that the ZM scheme is triggered, i.e., dilute CAPE exceeds $\geq 70$ J/kg. The regression indicates that model simulations with greater subsidence also have less convective activity.

@@

Figure~\ref{fig:4zonal}a,b, shows the zonal mean variations in $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ and $FREQZM$ in the convergence experiment. $FREQZM$ is largest in the $\pm 10^{\circ}$ latitude region, within the Intertropical Convergence Zone (ITCZ), and rapidly decreasing polewards into the subtropics. $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ increases away from the ITCZ region and reaches a maximum at the poleward limit of the Hadley Cell. The increase (decrease) in $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ ($FREQZM$) with resolution is to first-order, independent of latitude.

@@

The zonal mean R-values indicate the greatest goodness of fit in the $\pm 10^{\circ}$ latitude band, hereafter referred to as the deep tropics. In this region, the shape parameter $b_1$ is large and negative (Figure~\ref{fig:4zonal}d), consistent with the idea that subsiding motion stabilizes the environment and actively depresses dilute CAPE and the activity of the ZM scheme in the simulations. The shape parameter becomes less negative in the deep tropics with resolution, likely due to the greater magnitude $\langle f_{d} \rangle \langle \omega_{d} \rangle$ with resolution, which requires a lower $b_1$ to predict the binary $FREQZM$. The R-values generally decrease with resolution indicating that there is degradation in the relationship with resolution.

@@

Table~\ref{tbl:table2} shows the fractional contribution of the deep tropics to the climatological, global mean change in convective precipitation with resolution. The table indicates that a majority ($60-70 \%$) of the reduction in convective precipitation with resolution is from changes within the deep tropics (except in going from $ne20$ to $ne30$, where convective precipitation rates increase, in part due to a wide double-ITCZ in the $ne20$ run that spans outside of $\pm 10^{\circ}$ latitude, but which is contained within $\pm 10^{\circ}$ latitude in the $ne30$ run). Expanding the latitude boundaries marginally to $\pm 15^{\circ}$, roughly $75\%$ of the changes in convective precipitation with resolution occurs in this region (again, ignoring $ne30-ne20$; Table~\ref{tbl:table2}). This trend reflects changes in the partitioning of the ITCZ from convective to stratiform precipitation with resolution.

@@

The space-time weights associated with ascending and descending grid columns in the deep tropics vary drastically with resolution (Figure~\ref{fig:cape}b). The subsiding (ascending) space-time weights change from $0.32$ ($0.68$) in the $ne20$ run, monotonically increasing (decreasing) with resolution to $0.51$ ($0.49$) in the $ne120$ run. This increasing occurrence of stable, subsiding grid columns accounts for about half of the changes in dilute CAPE in the deep tropics with resolution, the other half being due to the systematic reduction in dilute CAPE for both ascending and descending regions (Figure~\ref{fig:cape}a). This is verified through comparing the weighted sum of the ascending/descending dilute CAPE values with the dilute CAPE for the entire deep tropics (compare black and grey crosses in Figure~\ref{fig:cape}a).

To isolate the relative importance of temperature or moisture on the systematic reduction in dilute CAPE with resolution for both ascending/descending regimes, dilute CAPE is recomputed for all resolutions, but through fixing the moisture profiles to the lowest resolution $ne20$ profile, and then through only fixing only the temperature to the $ne20$ profile. Figure~\ref{fig:cape}d shows the influence of changing temperature profile with resolution on dilute CAPE of ascending/descending grid columns, and illustrates that the systematic reduction in dilute CAPE with resolution in both regimes is primarily from changes to the temperature field. Moisture changes with resolution has a smaller influence, only appreciably impacting the dilute CAPE of ascending regions at higher resolutions (not shown).

@@

Figure~\ref{fig:cape-subt}a shows the dilute CAPE values computed from mean temperature and moisture of subsiding and ascending regions in the $\pm \left( 10^{\circ}-30^{\circ} \right) $ latitude bands, hereafter referred to as the subtropics. The spread in dilute CAPE between ascending and descending grid columns is much smaller than for the deep tropics (Figure~\ref{fig:cape}a), and their dilute CAPE  values vary much less with resolution ($\sim 20$ J/kg, compared with $\sim 80$ J/kg across all resolutions in the deep tropics). Through recomputing dilute CAPE and fixing the temperature or moisture profiles to $ne20$ values, the reduction in dilute CAPE with resolution is attributed to both temperature and moisture, but with temperature changes playing a larger role (not shown). This analysis suggests that Increasing subsidence with resolution increases the stability of the mean state, since both ascending/descending regimes become more stable with resolution, and reducing climatological $FREQZM$ despite the poor relationship between instantaneous subsidence and $FREQZM$ in the subtropics, as shown by the logistic regression (Figure~\ref{fig:4zonal}c).

@@

In contrast to the impact of vertical motion on the ZM scheme, the stratiform scheme is more intuitively connected to vertical velocities. \cite{RETAL2016CD} proposed an approximate scaling for the total precipitation rate in models $P_{tot}$, proportional to the upward moisture flux through cloud base,
\begin{equation}
P_{tot} \approx -\frac{1}{g\rho_{w}} \omega^{cb}_u q^{cb}_u \label{eq:mflux}
\end{equation}
where $\omega^{cb}_u$ and $q^{cb}_u$ refers to $\omega$ and specific humidity $q$ for ascending motion at cloud base, respectively, with $g=9.80616$ m/s$^2$ the acceleration of gravity and $\rho_w=1000$ kg/m$^3$ the density of rainwater used in CAM. \cite{OETAL2016JAMES} deem this the ``what goes up, must go down" model, highlighting the physically intuitive nature of the scaling, but note the scaling omits other potentially important processes, such as detrainment of cloudy updrafts into the environment or updrafts embedded within efficient circulations, with large gross moist stabilities.

Through approximating the cloud base as the $850$ hPa level, equation~\ref{eq:mflux} was found to provide a good fit to total precipitation rates in a regional model \citep{RETAL2016CD} and in the CAM-SE AGCM with CAM5 physics, across multiple resolutions \citep{OETAL2016JAMES}. As noted by \cite{RETAL2016CD}, using only resolved quantities in equation~\ref{eq:mflux} omits the sub-grid scale contribution to moisture fluxes, and so it is instructive to analyze the relative skill of this approximation for the individual components of $P_{tot}$, being the sum of stratiform $P_s$ and ZM convective $P_c$ precipitation rates. Figure~\ref{fig:mflux} shows the median $P_c$, $P_s$ and $P_{tot}$ conditioned on the $850$ hPa resolved moisture flux scaling using $1$ mm/day bins, in the CAM6 $ne30$ simulation. The figure shows that the moisture flux scaling is a good first-order approximation to the median total precipitation rate, and that the increase in precipitation rates with moisture flux is primarily from the stratiform precipitation scheme. The scaling is just as skillful for different resolutions \citep[not shown;][]{OETAL2016JAMES} and so changes to the global mean, climatological $P_s$ with resolution can be understood through quantifying changes in resolved moisture fluxes with resolution.

To unravel the contributions of changes in $\omega$ and $q$ at the 850 hPa level $\omega_{850}$ and $q_{850}$, to the increase in global mean, climatological stratiform precipitation rates $\overline{P_s}$, $\overline{P_s}$ is decomposed in $\omega_{850}$ and $q_{850}$ space following \cite{TETAL2018CD}. Table~\ref{tbl:table2} indicates that the $\pm 15^{\circ}$ latitude band accounts for most of the change in global mean stratiform precipitation with resolution. The decomposition is then restricted to stratiform precipitation rates and moisture fluxes in the $\pm 15^{\circ}$ latitude region to simplify the analysis, and $\overline{P_s}$ is redefined as the average over $\pm 15^{\circ}$ latitude.

Figure~\ref{fig:pdecomp} shows plots of the terms $M_s$ and $f_s M_s$ for all resolutions. The plots are computed using 6-hourly instantaneous output of $\omega_{850}$ and $q_{850}$, with 0.05 Pa/s and 0.4 g/kg bin wdiths, respectively. Bins in  $\left( \omega_{i} , q_{j} \right)$ space with $f_s <1 \times 10^{-5}$ are masked out, which is a somewhat arbitrary, but still reasonable cut-off for a bins' contribution to $\overline{P_s}$. The $M_s$ plots (left column) show that larger magnitude $\omega_{850}$ correspond to larger magnitude stratiform precipitation rates, while the impact of changing $q_{850}$ on $M_s$ is less clear. The changes in $M_s$ with resolution are subtle, while the changes in $f_s$ with resolution are large (not shown). The changes to $f_s$ can be inferred from the larger space of $\left( \omega_i , q_j \right)$ plotted at higher resolutions, indicating that larger magnitude $\omega_{850}$ values are occurring more frequently at higher resolutions, above the cutoff for plotting. The $f_s M_s$ plots (right column) clearly shows that larger magnitude $\omega_{850}$, and therefore larger magnitude stratiform precipitation rates are contributing to the increase in $\overline{P_s}$ with resolution. In contrast, smaller magnitude $\omega_{850}$ and hence smaller stratiform precipitation rates contribute less and less to $\overline{P_s}$ with increasing resolution.

>>>>

\cite{RETAL2016CD} proposed an approximate scaling for the total precipitation rate in models $P_{tot}$, proportional to the upward moisture flux through cloud base,
\begin{equation}
P_{tot} \approx -\frac{1}{g\rho_{w}} \omega q |_{cb} \label{eq:mflux}
\end{equation}
where $\omega q |_{cb}$ refers to the product of $\omega$ and specific humidity $q$ for ascending motion at cloud base, respectively, with $g=9.80616$ m/s$^2$ the acceleration of gravity and $\rho_w=1000$ kg/m$^3$ the density of rainwater used in CAM. \cite{OETAL2016JAMES} deem this the ``what goes up, must go down" model, highlighting the physically intuitive nature of the scaling, but note the scaling omits other potentially important processes, such as detrainment of cloudy updrafts into the environment or updrafts embedded within efficient circulations, with large gross moist stabilities.

Through approximating the cloud base as the $850$ hPa level, equation~\ref{eq:mflux} was found to provide a good fit to total precipitation rates in a regional model \citep{RETAL2016CD} and in the CAM-SE AGCM with CAM5 physics, across multiple resolutions \citep[recall that CAM-SE is different from CAM-SE-CSLAM, used in this study;][]{OETAL2016JAMES}. As noted by \cite{RETAL2016CD}, using only resolved quantities in equation~\ref{eq:mflux} omits the sub-grid scale contribution to moisture fluxes, and so it is instructive to analyze the relative skill of this approximation for the individual components of $P_{tot}$, being the sum of stratiform $P_s$ and parameterized deep convective $P_c$ precipitation rates. Figure~\ref{fig:mflux} shows the median $P_c$, $P_s$ and $P_{tot}$ conditioned on the $850$ hPa resolved moisture flux scaling using $1$ mm/day bins, in the CAM6 $ne30$ simulation. The figure shows that the moisture flux scaling is a good first-order approximation to the median total precipitation rate, and that the increase in precipitation rates with moisture flux is primarily from the stratiform precipitation scheme. The scaling is just as skillful for different resolutions \citep[not shown;][]{OETAL2016JAMES} and so changes in stratiform precipitation rate with resolution can be understood through quantifying changes in resolved moisture fluxes with resolution.

To unravel the contributions of changes in $\omega$ and $q$ at the 850 hPa level, $\omega_{850}$ and $q_{850}$, to the increase in global mean, climatological stratiform precipitation rates $\overline{P_s}$, $\overline{P_s}$ is decomposed in $\omega_{850}$ and $q_{850}$ space following \cite{TETAL2018CD}. Table~\ref{tbl:table2} shows the fractional contribution of deep tropical belts to the global mean change in stratiform precipitation with resolution. The table indicates that roughly a majority ($45-70 \%$) of the increase in stratiform precipitation with resolution is from changes within the $\pm 15^{\circ}$ latitude band, and so the moisture flux decomposition is restricted to this region to simplify the analysis; $\overline{P_s}$ is redefined as the average over $\pm 15^{\circ}$ latitude.

Figure~\ref{fig:pdecomp} shows plots of the terms $M_s$ and $f_s M_s$ for all resolutions. The plots are computed using 6-hourly instantaneous output of $\omega_{850}$ and $q_{850}$, with 0.05 Pa/s and 0.4 g/kg bin widths, respectively. Bins in  $\left( \omega_{i} , q_{j} \right)$ space with $f_s <1 \times 10^{-5}$ are masked out, which is a somewhat arbitrary, but still reasonable cut-off for a bins' contribution to $\overline{P_s}$. The $M_s$ plots (left column) show that larger magnitude $\omega_{850}$ correspond to larger magnitude stratiform precipitation rates, while the impact of changing $q_{850}$ on $M_s$ is less clear. The changes in $M_s$ with resolution are subtle, while the changes in $f_s$ with resolution are large (not shown). The changes to $f_s$ can be inferred from the larger space of $\left( \omega_i , q_j \right)$ plotted at higher resolutions, indicating that larger magnitude $\omega_{850}$ values are occurring more frequently at higher resolutions, above the cutoff for plotting. The $f_s M_s$ plots (right column) clearly shows that larger magnitude $\omega_{850}$, and therefore larger magnitude stratiform precipitation rates, are contributing to the increase in $\overline{P_s}$ with resolution. In contrast, smaller magnitude $\omega_{850}$ and hence smaller stratiform precipitation rates contribute less and less to $\overline{P_s}$ with increasing resolution.

@@

While the increase in magnitude of vertical velocities with resolution clearly drives the increase in moisture fluxes and stratiform precipitation rates with resolution, it is not clear why there is a reduction in parameterized deep convective precipitation rates with resolution. Curiously though, there is an excellent negative correlation (Pearson's R-value = 0.99, N = 27) between the global mean, climatological $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ and a measure of the activity of the \cite{ZM1995AO} deep convection scheme (referred to as the {\em{ZM scheme}} hereafter), global mean, climatological $FREQZM$ (Figure~\ref{fig:corr}). At any grid-point and time-step, $FREQZM$ is a binary variable: 1 if the ZM scheme is active, 0 if it is not. Time mean $FREQZM$ is therefore the fraction of the model time that the ZM scheme is triggered, i.e., dilute CAPE exceeds $\geq 70$ J/kg. The regression indicates that model simulations with greater subsidence also have less convective activity.

@@

Figure~\ref{fig:4zonal}a,b, shows the zonal mean variations in $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ and $FREQZM$ in the convergence experiment. $FREQZM$ is largest in the $\pm 10^{\circ}$ latitude region, within the ITCZ, and rapidly decreasing polewards into the subtropics. $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ increases away from the ITCZ region and reaches a maximum at the poleward limit of the Hadley Cell. The increase (decrease) in $\langle f_{d} \rangle \, \langle \omega_{d} \rangle$ ($FREQZM$) with resolution is to first-order, independent of latitude.

@@

The zonal mean R-values indicate the greatest goodness of fit in the $\pm 10^{\circ}$ latitude band, hereafter referred to as the deep tropics. In this region, the shape parameter $b_1$ is large and negative (Figure~\ref{fig:4zonal}d), consistent with the idea that subsiding motion stabilizes the environment and actively depresses dilute CAPE and the activity of the ZM scheme in the simulations. The shape parameter becomes less negative in the deep tropics with resolution, likely due to the greater magnitude $\langle f_{d} \rangle \langle \omega_{d} \rangle$ with resolution, which requires a lower $b_1$ to predict the binary $FREQZM$. The R-values generally decrease with resolution indicating that there is degradation in the relationship with resolution. The logistic regression is less skillful in the subtropics, with skill declining even further in the mid-latitudes (Figure~\ref{fig:4zonal}c).

@@

Table~\ref{tbl:table2} indicates that a majority ($60-70 \%$) of the reduction in convective precipitation with resolution is from changes within the deep tropics (except in going from $ne20$ to $ne30$, where convective precipitation rates increase, in part due to a wide double-ITCZ in the $ne20$ run that spans outside of $\pm 10^{\circ}$ latitude, but which is contained within $\pm 10^{\circ}$ latitude in the $ne30$ run). Expanding the latitude boundaries marginally to $\pm 15^{\circ}$, roughly $75\%$ of the changes in convective precipitation with resolution occurs in this region (again, ignoring $ne30-ne20$; Table~\ref{tbl:table2}). This trend reflects changes in the partitioning of the ITCZ from convective to stratiform precipitation with resolution.

@@

The space-time weights associated with ascending and descending grid columns in the deep tropics vary drastically with resolution (Figure~\ref{fig:cape}b). The subsiding (ascending) space-time weights change from $0.32$ ($0.68$) in the $ne20$ run, monotonically increasing (decreasing) with resolution to $0.51$ ($0.49$) in the $ne120$ run. The increasing occurrence of stable, subsiding grid columns accounts for about half of the changes in dilute CAPE in the deep tropics with resolution, the other half being due to the systematic reduction in dilute CAPE for both ascending and descending regions (Figure~\ref{fig:cape}a). This breakdown is straightforward to compute since the sum of the products of the space-time fractions with their associated ascending/descending dilute CAPE values approximate the dilute CAPE computed from the mean temperature/moisture profiles over the entire deep tropics (compare black and grey crosses in Figure~\ref{fig:cape}a) rather well.

To isolate the relative importance of temperature or moisture on the systematic reduction in dilute CAPE with resolution for both ascending/descending regimes, dilute CAPE is recomputed for all resolutions, but through fixing the moisture profiles to the lowest resolution $ne20$ profile, and then through fixing only the temperature to the $ne20$ profile. Figure~\ref{fig:cape}d shows the influence of changing temperature profile with resolution on dilute CAPE of ascending/descending grid columns, and illustrates that the systematic reduction in dilute CAPE with resolution in both regimes is primarily from changes to the temperature field. Moisture changes with resolution has a smaller influence, only appreciably impacting the dilute CAPE of ascending regions at higher resolutions (not shown).

@@

Figure~\ref{fig:cape-subt}a shows the dilute CAPE values computed from mean temperature and moisture of subsiding and ascending regions in the $\pm \left( 10^{\circ}-30^{\circ} \right) $ latitude bands, hereafter referred to as the subtropics. The spread in dilute CAPE between ascending and descending grid columns is much smaller than for the deep tropics (Figure~\ref{fig:cape}a), and their dilute CAPE  values vary much less with resolution ($\sim 20$ J/kg, compared with $\sim 80$ J/kg across all resolutions in the deep tropics). Through recomputing dilute CAPE and fixing the temperature or moisture profiles to $ne20$ values, the reduction in dilute CAPE with resolution is attributed to both temperature and moisture, but with temperature changes playing a larger role (not shown). This analysis suggests that increasing subsidence with resolution (\ref{fig:4zonal}a) increases the stability of the mean state, since both ascending/descending regimes become more stable with resolution, and reducing climatological $FREQZM$ (\ref{fig:4zonal}b) despite the poor relationship between instantaneous subsidence and $FREQZM$ in the subtropics, as shown by the logistic regression (Figure~\ref{fig:4zonal}c).

@@
